\sffamily An EEG-based visual Brain-Computer Interface for gaze-free communication

### 2024-09-28 chapters/stbf_struct.tex
started wcble introduction
Comments François, smearing figure

	Part of the work in this chapter was published in~\cite{VanDenKerchove2022}.

@@

	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when $\hat{C}_\text{emp}$ is singular.
	\autoref{fig:kronlda-covs}a and \autoref{fig:covs}b respectively show examples of the
	empirical estimators of the covariance and the inverse covariance matrices.
	The empirical estimator suffers from performance and
	stability issues if the number of epochs $n$ used or estimation is not much larger than the number of features $cs$~\cite{Stein1956,Khatri1987}.

@@

	with $0<\alpha<1$.
	Analogous to L2 regularization of the beamforming problem,
	shrinkage reduces the ratio between the smallest and largest eigenvalues
	of the covariance matrix by strengthening the diagonal.
	\autoref{fig:kronlda-covs}c and \autoref{fig:covs}d respectively show examples of the
	shrunk estimator of the covariance and the inverse covariance matrices.

@@

	Using \autoref{eq:closed_form_kronecker} removes the need to calculate the full, high dimensional Kronecker product $\hat{S}^+\otimes
		\hat{T}^+$.
	\autoref{fig:kronlda-covs}e and \autoref{fig:covs}f respectively show examples of the
	structured covariance and inverse covariance estimators,
	consisting of a spatial Kronecker factor (\autoref{fig:kronlda-covs}g and
	\autoref{fig:kronlda-covs}h) and a temporal component (\autoref{fig:covs}i and
	\autoref{fig:kronlda-covs}j).

@@

	The Kronecker approach has shown significant performance yields in different linear spatiotemporal EEG and MEG
	applications~\cite{DeMunck2002,Huizenga2002,6408231,GonzalezNavarro2016,GonzalezNavarro2017}.
	Van Vliet \& Salmelin~\cite{Vliet2020} have applied a Kronecker-structured covariance estimator to ERP classification with linear models in a post-hoc fashion.
	Our work goes further by embedding the Kronecker structure in the
	spatiotemporal beamformer training process, using a data-adaptive shrinkage
	method, and regularizing the covariance further by imposing a Toeplitz
	structure on the temporal covariance.

	\subsubsection{Kronecker-Toeplitz structured covariance estimation}
	\label{sec:structured_covariance}
	The question remains how to estimate $\hat{S}$ and $\hat{T}$.
	While the Flip-Flop and Non-iterative Flip-Flop
	algorithms~\cite{Lu2005, Werner2008, Wirfaelt2010} can estimate Kronecker or Kronecker-Toeplitz structured covariances, new results show that a fixed point iteration is more efficient~\cite{Wiesel2012a,Wiesel2012}.
	After each iteration, the spatial and temporal covariances matrices are scaled to unit
	variance to ensure the fixed point iteration converges.
	Finally, shrinkage can also be introduced in the Fixed Point Iteration to
	improve stability and achieve more robust
	regularization~\cite{Wiesel2012,Greenewald2014,6408231, Breloy2016}.
	The spatial and temporal covariance matrices are shrunk at every fixed-point
	iteration with shrinkage factors $\beta_k$ and $\gamma_k$ before matrix
	inversion in the
	next iteration.
	Combined, this leads to the iterative estimation algorithm described by the
	following equations:

@@

	\section{Results}
	\subsection{Minimum required fixed-point iterations}
	The fixed point iteration algorithm described
	in Equations \ref{eq:fpi_spatial}-\ref{eq:fpi_temporal_shrunk} estimates the Kronecker-Toeplitz structured covariance for the
	\textsc{stbf-struct} classifier.
	Fixed-point iteration is an iterative procedure starting from (in our case)
	non-informed initial guesses for the spatial and temporal covariance matrices.
	As a stopping criterion, one could impose a threshold on the difference in
	outcome of successive steps, e.g., based on the covariance norm or the
	classifier accuracy.
	However, few iterations or even just one~\cite{castaneda2014estimation} suffice to achieve satisfactory performance in practice.

@@

	\subsection{Classifier accuracy for limited training data}
	It is of interest to keep the calibration time before BCI
	operation as short as possible.
	We mimic this problem by training the classifier with as few training epochs as possible.
	We evaluate the performance of all classifiers for different levels of
	available training data and apply the cross-validation procedure nine times (the number of blocks in the training fold) for all subjects, keeping the
	corresponding number of blocks in the training folds and dropping the rest.
	\autoref{fig:accuracy} and \autoref{ap:accuracy_ap} show each classifier's
	accuracy relative to the data availability.
	We statistically compare the two newly proposed classifiers,
	\textsc{stbf-struct} and \textsc{stbf-shrunk}, for different levels of training
	data availability using a one-sided paired Wilcoxon rank-sum test with Holm correction for the multiple pairwise comparisons between classifiers.
	We performed this analysis three times: by only using the first trial of a
	block, by averaging epochs across the first two trials of a block and across
	the first five trials of a block.
	Results validated on one trial are reported in \autoref{tab:p_values_1}, 2-trial results
	in \autoref{tab:p_values_2}, and 5-trial results in \autoref{tab:p_values_5}.

  \begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/accuracy.eps}
		\caption{Accuracy of the different classifiers for all 21 subjects relative to the
			number of blocks available for training. One block consists of 135
			epochs and corresponds to 27 seconds of stimulation. Accuracies
			are shown for the evaluation settings averaging over 1, 2, and
			3 trials of testing stimuli.
			\autoref{ap:accuracy_ap} contains results for all numbers of trials.
			While \textsc{stbf-emp} is unstable when little training data
			are available,
			regularization of the covariance matrix (\textsc{stbf-shrunk} and
			\textsc{stbf-struct}) drastically improves performance.}
		\label{fig:accuracy}
	\end{figure}
	\begin{table}[h]
		\centering
		\input{chapters/stbf_struct_tests_1.tex}
		\caption{$p$-values calculated by one-sided paired Wilcoxon rank-sum test with
			Holm correction using one testing trial for different classifiers
			and levels of data availability. $p$-values $<0.05$ are considered significant
			and marked in grey.}
		\label{tab:p_values_1}

@@

>>>>

  \emph{This chapter, except section~\ref{sec:stbf-struct-kronlda}, was
  published by~\textcite{VanDenKerchove2022}}.

@@

	The Moore-Penrose pseudoinverse $^+$, ensures a solution exists when $\hat{C}_\text{emp}$ is singular.
	\autoref{fig:kronlda-covs}a and \autoref{fig:kronlda-covs}b respectively show examples of the
	empirical estimators of the covariance and the inverse covariance matrices.
	The empirical estimator suffers from performance and
	stability issues if the number of epochs $n$ used or estimation is not much larger than the number of features $cs$~\cite{Stein1956,Khatri1987}.

@@

	with $0<\alpha<1$.
	Analogous to L2 regularization of the beamforming problem,
	shrinkage reduces the ratio between the smallest and largest eigenvalues
	of the covariance matrix by strengthening the diagonal.
	\autoref{fig:kronlda-covs}c and \autoref{fig:kronlda-covs}d respectively show examples of the
	shrunk estimator of the covariance and the inverse covariance matrices.

@@

	Using \autoref{eq:closed_form_kronecker} removes the need to calculate the full, high dimensional Kronecker product $\hat{S}^+\otimes
		\hat{T}^+$.
	\autoref{fig:kronlda-covs}e and \autoref{fig:covs}f respectively show examples of the
	structured covariance and inverse covariance estimators,
	consisting of a spatial Kronecker factor (\autoref{fig:kronlda-covs}g and
	\autoref{fig:kronlda-covs}h) and a temporal component (\autoref{fig:kronlda-covs}i and
	\autoref{fig:kronlda-covs}j).

@@

	The Kronecker approach has shown significant performance yields in different linear spatiotemporal EEG and MEG
	applications~\cite{DeMunck2002,Huizenga2002,Beltrachini2013,GonzalezNavarro2016,GonzalezNavarro2017}.
	Van Vliet \& Salmelin~\cite{Vliet2020} have applied a Kronecker-structured covariance estimator to ERP classification with linear models in a post-hoc fashion.
	Our work goes further by embedding the Kronecker structure in the
	spatiotemporal beamformer training process, using a data-adaptive shrinkage
	method, and regularizing the covariance further by imposing a Toeplitz
	structure on the temporal covariance.

	\subsubsection{Kronecker-Toeplitz structured covariance estimation}
	\label{sec:structured_covariance}
	The question remains how to estimate $\hat{S}$ and $\hat{T}$.
	While the Flip-Flop and Non-iterative Flip-Flop
	algorithms~\cite{Lu2005, Werner2008, Wirfaelt2010} can estimate Kronecker or Kronecker-Toeplitz structured covariances, new results show that a fixed point iteration is more efficient~\cite{Wiesel2012a,Wiesel2012}.
	After each iteration, the spatial and temporal covariances matrices are scaled to unit
	variance to ensure the fixed point iteration converges.
	Finally, shrinkage can also be introduced in the Fixed Point Iteration to
	improve stability and achieve more robust
	regularization~\cite{Wiesel2012,Greenewald2014,Beltrachini2013, Breloy2016}.
	The spatial and temporal covariance matrices are shrunk at every fixed-point
	iteration with shrinkage factors $\beta_k$ and $\gamma_k$ before matrix
	inversion in the
	next iteration.
	Combined, this leads to the iterative estimation algorithm described by the
	following equations:

@@

	\section{Results}
	\subsection{Minimum required fixed-point iterations}
	The fixed point iteration algorithm described
	in Equations \ref{eq:fpi_spatial}-\ref{eq:fpi_temporal_shrunk} estimates the Kronecker-Toeplitz structured covariance for the
	\textsc{stbf-struct} classifier.
	Fixed-point iteration is an iterative procedure starting from (in our case)
	non-informed initial guesses for the spatial and temporal covariance matrices.
	As a stopping criterion, one could impose a threshold on the difference in
	outcome of successive steps, e.g., based on the covariance norm or the
	classifier accuracy.
	However, few iterations or even just one~\cite{Castaneda2014} suffice to achieve satisfactory performance in practice.

@@

	\subsection{Classifier accuracy for limited training data}
	It is of interest to keep the calibration time before BCI
	operation as short as possible.
	We mimic this problem by training the classifier with as few training epochs as possible.
	We evaluate the performance of all classifiers for different levels of
	available training data and apply the cross-validation procedure nine times (the number of blocks in the training fold) for all subjects, keeping the
	corresponding number of blocks in the training folds and dropping the rest.
	\autoref{fig:accuracy} and \autoref{fig:accuracy_ap} show each classifier's
	accuracy relative to the data availability.
	We statistically compare the two newly proposed classifiers,
	\textsc{stbf-struct} and \textsc{stbf-shrunk}, for different levels of training
	data availability using a one-sided paired Wilcoxon rank-sum test with Holm correction for the multiple pairwise comparisons between classifiers.
	We performed this analysis three times: by only using the first trial of a
	block, by averaging epochs across the first two trials of a block and across
	the first five trials of a block.
	Results validated on one trial are reported in \autoref{tab:p_values_1}, 2-trial results
	in \autoref{tab:p_values_2}, and 5-trial results in \autoref{tab:p_values_5}.

  \begin{figure}
		\includegraphics[width=\linewidth]{figures/stbf_struct/accuracy.eps}
		\caption{Accuracy of the different classifiers for all 21 subjects relative to the
			number of blocks available for training. One block consists of 135
			epochs and corresponds to 27 seconds of stimulation. Accuracies
			are shown for the evaluation settings averaging over 1, 2, and
			3 trials of testing stimuli.
			\autoref{fig:accuracy_ap} contains results for all numbers of trials.
			While \textsc{stbf-emp} is unstable when little training data
			are available,
			regularization of the covariance matrix (\textsc{stbf-shrunk} and
			\textsc{stbf-struct}) drastically improves performance.}
		\label{fig:accuracy}
	\end{figure}
	\begin{table}[h]
		\centering
		\input{chapters/stbf_struct_tests_1.tex}
		\caption{$p$-values calculated by one-sided paired Wilcoxon rank-sum test with
			Holm correction using one testing trial for different classifiers
			and levels of data availability. $p$-values $<0.05$ are considered significant
			and marked in grey.}
		\label{tab:p_values_1}

@@

  \subsection{Extension to \ac{lda}}
  \label{sec:stbf-struct-kronlda}
  \todo{cite proof (or reiterate)}

### 2024-09-28 chapters/wcble.tex
Comments François, smearing figure

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli.
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform.\todomvh{refer to image in bci chapter}
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}.\todo{cite}
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for for gaze-impaired patients.

@@

The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
averaged per condition to cope with the unfavorable \ac{snr}?
When averaging over multiple identical responses with jittered latencies, the shape
and amplitude of the average do not reflect the properties
of the original signal.
Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting \ac{erp}
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{Woody1967, Pfefferbaum1980, McDowell2003, Verleger2005,
	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
	these references}.

@@

The most straight-forward method to estimate single-trial latency is
\emph{peak picking}
Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude relative to stimulus
onset.
While straight-forward, this method does not perform well in low
\ac{snr} conditions, unless combined with filtering.
Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}.
Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{
  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999,  Sparacino2002,
  Georgiadis2005,	D’Avanzo2011}
and in the time-frequency domain~\cite{
  Quiroga2003, Wang2007,Hu2010}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.

@@

\subsection{Multi-component approaches}
\label{sec:wcble/literature/multi-comp}
In the previous section, we have abstracted away the fact that \acp{erp} are
multi-component signals, which is not taking into account by the aforementioned
methods.
Multiple components can be time-locked to different events or neurophysiological
proccess and therefore are not necessarily time-locked to each other. They
each can have different latency variabilities.
The presence of more than one component can hamper the
performance of some latency estimation algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple \ac{erp} components
by carefully selecting peaks or pre-determined regions of interest.
\cite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
time-window for each \ac{erp} component to extract a latency per component.
Methods based on spatial decomposition filters like \ac{pca}, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
\ac{erp} components. A clear evaluation is lacking of the performance of these
methods in
multi-component settings is lacking.

@@

\section{Discussion \& conclusion}
In this work, we introduced a new \ac{erp} latency estimation and decoding
method that applies a spatiotemporal classification iteratively along the time
dimension of the data, refining the dataset at each iteration by aligning
trials.
The algorithm combines \acf{cble} with Woody's template matching iteration
scheme and is therefore named \acf{wcble}.
Our method attempts to model the spatial and temporal activation of
each \ac{erp} component, as well as it's latency distribution.
Through the firs-stage classifier, it takes into acount the
spatial and temporal structure of EEG background noise and the trials from the
classes to discriminate.
\section{Conclusion}

>>>>

\Ac{eeg}-based visual oddball \acp{bci} establish communication with paralyzed
individuals by decoding \acp{erp} obtained from time-modulated flashing
stimuli.
These \acp{erp} consist of multiple \ac{erp}-components, peaks and troughs in
\ac{erp} waveform.\todomvh{refer to image in bci chapter}
These components each have their own distinct amplitude and latency.
\Ac{erp} component latency can reflect a wide range of cognitive,
neurophysiological and clinical properties, that can be confounding factors in
the development of \acp{bci}.\todo{cite}
Latency is often assessed as the timing of an \ac{erp} component peak, obtained
after averaging over trials.
Latencies of non-averaged, single-trial \acp{erp} should capture even more
relevant information, yet they are very hard to measure due to low \ac{snr} of
single trial \acp{erp}.
Within-session variability of \ac{erp} latency, called \ac{erp} latency
\emph{jitter}, has been shown to influence \ac{bci}
accuracy~\cite{Thompson2012}.
This effect is especially prominent when bringing \ac{bci} development from the
lab to the clinical setting with patients in the target population.
\textcite{Zisk2021} showed that patients with \ac{als} have significantly
increased ERP jitter.
\textcite{Arico2014} show that latency jitter negatively affects \ac{bci}
performance, and disproportionally so when the user is not directly gazing at
intended targets, such as is the case for gaze-impaired patients.

@@

The presence of latency jitter introduces multiple issues to \ac{erp} analysis, the
most prominent of which is the smearing effect as illustrated
in~\autoref{fig:wcble/smearing}. In \ac{erp} analysis, single-trial \acp{erp} are usually
averaged per condition to cope with the unfavorable \ac{snr}?
When averaging over multiple identical responses with jittered latencies, the shape
and amplitude of the average do not reflect the properties
of the original signal.
Because the amplitude of the average is lower, the \ac{snr} will also be negatively impacted.
More generally, the smearing effect leads to difficulties interpreting \ac{erp}
results and making inferences about amplitude or latency effects, because these
are entangled when using averaging~\cite{Woody1967, Pfefferbaum1980, McDowell2003, Verleger2005,
	Roth2007, Walhovd2008, Poli2010, Luck2014}\todo{Review relevance of
	these references}.

\begin{figure}[t]
  \centering
  \input{figures/wcble/smearing.tikz.tex}
  \caption{The smearing effect. When obtaining an \ac{erp} template through
  averaging, high jitter causes a decrease in peak amplitude and a deformation
  of the shape of the waveform.}
  \label{fig:wcble/smearing}
\end{figure}

@@

The simplest method to estimate single-trial latency is
\emph{peak picking}
Here, the latency of an \ac{erp} is determined as the
time-point of it's maximum (or minimum) amplitude relative to stimulus
onset.
While straightforward, this method does not perform well in low
\ac{snr} conditions, unless combined with filtering.
Filtering can suppress the noise contaminating \ac{eeg} trials, lowering
the risk of picking a noise peak instead of a true peak of the \ac{erp}.
Several filtering approaches have been developed to be used in
conjuction with peak-picking, in the time domain~\cite{
  Carlton1980, McCarthy1981, Karjalainen1999, Nishida1999,  Sparacino2002,
  Georgiadis2005,	D’Avanzo2011}
and in the time-frequency domain~\cite{
  Quiroga2003, Wang2007,Hu2010}.
The aforementioned analysis and performance prediction method proposed
by \cite{Arico2014} estimates the single-trial P3 latencies of attended
epochs by P3 peak-picking after filtering in the time-frequency domain, and
corrects for jitter by temporally aligning all target epochs to the average
latency, i.e., by shifting the epochs in time such that the P3 peaks all fall
at the same time instance.
Filtering can
also be done in the spatial domain. The spatial filter can be constructed
using \ac{pca}~\cite{Ouyang2017}, ICA~\cite{Townsend1999, Jung2001,
	Milne2011}, or as a spatial LCMV beamformer~\cite{Treder2016}.

@@

\subsection{Multi-component approaches}
\label{sec:wcble/literature/multi-comp}
In the previous section, we have abstracted away the fact that \acp{erp} are
multi-component signals, which is not taking into account by the aforementioned
methods.
Multiple components can be time-locked to different events or neurophysiological
processes and therefore are not necessarily time-locked to each other. They
each can have different latency variabilities.
The presence of more than one component can hamper the
performance of some latency estimation algorithms, and it often only possible to extract only the
latencies of the largest component present~\cite{Ouyang2017}. When aligning trials
to one component and averaging, this will introduce blurring in the other
components~\cite{Ouyang2020}.

Most of the algorithms above can be adapted to work on multiple \ac{erp} components
by carefully selecting peaks or pre-determined regions of interest.
\cite{Hardiansyah2020} uses an \ac{snr} boosting method, applied to a specific
time-window for each \ac{erp} component to extract a latency per component.
Methods based on spatial decomposition filters like \ac{pca}, ICA or XDAWN can be leveraged
to utilize multiple of their components as filters, resulting in multiple time
series where peack picking or Woody's algorithm can be applied. In general,
however, these algorithms lack an integrated approach to deal with multiple
\ac{erp} components. A clear evaluation of the performance of these
methods in multi-component settings is lacking.

@@

\section{Discussion \& conclusion}
In this work, we introduced a new \ac{erp} latency estimation and decoding
method that applies a spatiotemporal classification iteratively along the time
dimension of the data, refining the dataset at each iteration by aligning
trials.
The algorithm combines \acf{cble} with Woody's template matching iteration
scheme and is therefore named \acf{wcble}.
Our method attempts to model the spatial and temporal activation of
each \ac{erp} component, as well as it's latency distribution.
Through the firs-stage classifier, it takes into acount the
spatial and temporal structure of EEG background noise and the trials from the
classes to discriminate.
